Leaf traits drive differences in biomass partitioning among major plant functional types

### 2015-01-06 manuscript.Rnw
Much writing. This is a full draft!

@@

\textbf{
The partitioning of biomass into short-lived leaves and long-lived stems is one of the most uncertain  and influential components of  global vegetation models (GVMs), affecting the lifetime of carbon in ecosystems \cite{friend2014}, predictions of global carbon storage \cite{ise2010}, and the CO\textsubscript{2} response of net primary production \cite{dekauwe2014}. Although GVMs typically assume that the major woody plant functional types (PFTs) differ in biomass partitioning, empirical data have been lacking. Using a new global database on biomass partitioning from over 175 studies, we here show a strong effect of PFT on partitioning between stems and leaves. Contrary to prior expectation, we show that major woody PFTs do not differ in the amount of leaf area supported, but rather the amount of leaf mass per unit above ground biomass. Moreover, we find that these differences relate to systematic shifts in leaf mass per area, a leading trait within the leaf economics spectrum. Combined, these results uncover fundamental rules in how plants are constructed and allow for systematic benchmarking of biomass partitioning routines in current GVMs.
}

@@

\begin{itemize}

\item GVMs (including dynamic GVMs, which model the distribution of vegetation types, and those who use a prescribed vegetation distributions) are used to predict the trajectory of the global vegetation carbon sink in a changing climate. Biomass partitioning (related to, but not to be confused with allocation of net primary production) is important because short-lived leaves re-enter the decomposition cycle quickly, releasing much of its carbon back to the atmosphere. Long-lived stems act as a very long carbon storage term \cite{luyssaert2008}. As a result, partitioning affects the overall lifetime of carbon in the ecosystem, which has been shown to be an important determinant of overall carbon storage projections \cite{friend2014}.
\item DGVMs differ widely in how partitioning is calculated (de Kauwe) (though it should be mentioned that partitioning and allocation are not the same! - see Methods S2). As the global diversity in plants is typically simplified in a handful of plant functional types, it is common to assume that biomass partitioning indeed varies predictably between PFTs, as a basis for model parameterization. However, it is largely unknown whether PFTs differ systematically in BP. This is largely because an appropriate database has been lacking. 
\item It is our intent to evaluate and critique alternative models of biomass allocation (wbe etc), but rather to establish, empirically, patterns in biomass partitioning across plant functional types and plant size.
\item What is known is that gymnosperms have more leaf biomass per unit above-ground biomass (the 'leaf mass fraction', LMF) than angiosperms \cite{oneilldeangelis1981,enquistniklas2002}. But it is not even clear whether this is due to the angiosperm/evergreen divergence, or related to inherent differences in leaf life span. If the latter, we would expect evergreen angiosperms to have more leaf biomass than deciduous angiosperms. 
\item Novelty in this study, a) individual plants, b) entire size range including saplings and the largest trees in the world, c) estimates of leaf area as well as mass d) further division into three plant functional types, and three biomes
\item It is unclear whether PFTs with higher leaf biomass will have higher total leaf area, because PFTs differ in the ratio of leaf mass to area (LMA), a measure of the leaf construction cost. Leaf area itself is relevant because light interception acts on leaf area, which drives photosynthesis. A few classic studies have assumed that gymnosperms indeed carry much more leaf area than angiosperms (Bond; Chabot and Hicks). 
\item Finally it is also unknown whether potential differences between PFTs in BP are a result of inherent differences in morphology, or whether climate has direct effects on BP, and the PFTs simply occupy different climate spaces. 
\item Here we use BAAD, large database blah blah, three main PFTs, many biomes, wide range in plant size, good coverage of climate space where woody plants occur (Fig 1a and S1). We included several key variables from BAAD (Fig 1b). Crucially, the data are for individuals, which is more relevant to DGVMs than plot-level estimates, because ...
\item We hypothesize that gymnosperms have higher LMF than angiosperms, and that evergreen angiosperms have higher LMF than deciduous angiosperms (we did not have sufficient data to test deciduous evergreens). 
\item We found that LMF differed strongly between the three PFTs tested, with evergreen gymnosperms > evergreen angiosperms > deciduous angiosperms (Fig 2a). Because of allometric relationships between the studied variables (niklas), there was a very strong effect of plant height on LMF (Fig 2a and S2). We corrected for plant height by estimating LMF at a common plant height, using a generalized additive model, a semi-parametric method that makes no assumption on the underlying functional relationships. 
\item We then found that LMF was proportional to the leaf construction cost, as measured by leaf mass per area (Fig 2b). \item The consequence of this is that the leaf area ratio (the amount of leaf area supported per unit above-ground biomass), was invariant between PFTs, because $LAR = LMF / LMA$ (Fig 2c and S3). We thus supported the view that PFTs differ in LMF, but for the first time demonstrated that they converge in the amount of leaf area supported.
\item Next, we split LMF into two components, $LMF = M_F/M_T = M_F/A_S * A_S/MT$, where $M_F/A_S$ is the ratio of leaf biomass to stem basal area (see Methods S1), and $A_S/M_T$ the ratio of stem basal area to total above-ground biomass. We found that the first term showed identical PFT patterns as LMF (Fig 2d and e), $M_F/A_S$ was highly variable between PFT and proportional to LMA, so that $A_F/A_S$ is invariant between PFTs. 
\item This split is also convenient because the first term is nearly independent of plant height, allowing a more straightforward analysis of climate effects (not having to correct for height every time). The second term was found to be nearly unaffected by PFT (S4), but strongly height-dependent. It thus absorbs variation related to height and above-ground plant geometry, and is consistent with previous findings that this ratio is fairly conserved between plant species (Chaves).
\item We analyzed climate effects in two ways, biome, quantitative variables (MAP,MAT), and by biome (tropical / temperate/ boreal), which is a simple classification taking into account both MAT and MAP.
\item Regardless of how we analyzed it, we found only weak climate effects. We did not find differences between biomes in $M_F/A_S$ (Fig 3), but some small inconsistent differences in $A_F/A_S$ (corresponding to slightly lower LMA in tropical vs. temperate evergreen angiosperms). Biome consistently explained very little variation when added to a statistical model in addition to PFT and plant height (Table S6). 
\item We found statistically significant but weak and inconsistent effects of climate variables (MAP and MAT) on $M_F/A_S$ and $A_F/A_S$ (Fig 4). These variables explained very little variation when added to a statistical model in addition to PFT and plant height (Table 1). 
\item We found that PFTs differed in LMF, and this was proportional to leaf mass per area, a leading trait within the leaf economics spectrum. Indeed, if LMA was used instead of PFT, we were able to explain more variation in $M_F/A_S$ and $A_F/A_S$ (Table S7). It thus seems possible to integrate a key leaf trait with whole-plant modelling of biomass partitioning. 
\item We found very modest effects of climate on biomass partitioning, somewhat inconsistent with Reich et al 2014. It is consistent with the idea that LMA is more flexible than biomass partitioning (Poorter). We argue that climate directly or indirectly affects LMA, which then leads to differences in LMF, because LAR is less variable. Plants seem to target a certain total leaf area display per unit above-ground biomass investment, but we do stress that there is still much variation unaccounted for (Figure S histograms), and that other studies certainly have uncovered small-scale effects of climate on leaf area to sapwood area (pine studies etc), an important component of $A_F/A_S$. Our results hold at the PFT level, which is relevant for DGVMs.
\item What consequences for DGVMs? There is a multitude of allocation models (de Kauwe). Together with assumptions on leaf and stem turnover rates, this gives estimates of biomass partitioning (S2 methods). So far it is common to choose between different models largely based on philosophical arguments \cite{franklin2012}. We argue that relationships such as in Fig 2a should be used to benchmark and constrain the various allocation models. The data we use here is freely available (Falster), which should aid in the models being tested, with the hope that they perhaps converge.

\end{itemize}

\subsubsection*{Old text bit still useful}

Major plant functional types differ widely in leaf mass per area (LMA, the ratio of leaf mass to area, an index of leaf thickness) (\cite{poorter2009}), which can be expected to affect biomass partitioning in several ways. First, LMA correlates with leaf lifespan. Longer lived leaves would lead to higher plant leaf mass ($M_F$) as long as annual production rates are similar (\cite{enquistniklas2002}). Another interpretation is that plants may target a certain leaf area display given a certain biomass investment in supporting tissues, because light interception and photosynthesis correlate with the exchange area of the plant.

It has been reported numerous times that gymnosperms (with higher LMA) have higher leaf mass fraction than angiosperms (with lower LMA) \cite{enquistniklas2002, cheng2007a, mccarthyetal2007, poorter2012, sardans2013}, although a recent study reported no difference (\cite{wolf2010}). No study to date has been able to further tease apart PFT differences, or test the potentially confounding effects of climate.

It is customary to estimate total above-ground biomass from stem diameter \cite{jenkins2003,chave2005}.

We can further decompose LMF into two components representing different aspects of plant function and structure (see Methods). The ratio of leaf biomass ($M_F$) to stem basal area ($A_S$) relates to hydraulic supply, and the ratio of total above-ground biomass ($M_T$) and $A_S$ is a function of plant shape (stem taper, plant height at given stem diameter), and wood density \cite{chave2005}. We found that ratio of above-ground biomass to basal stem area ($M_T/A_S$) was nearly constant between PFTs (Fig.~\ref{fig:figureS4}), with deciduous angiosperms having slightly higher $M_T$ at a given $A_S$. However we found very large differences between PFTs in the ratio $M_F/A_S$(Fig.~\ref{fig:figure2}).

We then expressed the average $M_F/A_S$ as a function of the leaf mass per area (LMA = $M_F / A_F$), a measure of the leaf construction cost to the plant. We found that, across PFTs, $M_F/A_S$ was approximately proportional to the average leaf mass per area for the PFT. As a result of this approximate proportionality, the ratio of leaf area to basal stem area ($A_F/A_S$) was not different between PFTs (Fig.~\ref{fig:figure2}) (see also Fig.~\ref{fig:figureS5}). This relationship also held approximately when data were first averaged by species (Fig.~\ref{fig:figureS6}).

Unlike \cite{reich2014}, we found only very weak effects of climate on biomass relationships. The variance explained by MAT was very small, especially compared to the large difference between PFTs.

Leaf mass per area generally shows a more plastic response to climate than biomass partitioning  \cite{poorter2012}.

In summary, we found coordination between an important leaf trait and whole-plant biomass partitioning. This has implications for global vegetation models.

How is allocation implemented in current DGVMs? Either constant allocation fraction, target allometric equations, or optimization \cite{dekauwe2014}. Constant allocation fraction gives constant LAR by pft if leaf lifespan is exactly proportional to LMA. Target allometric equations can be made pft-invariant by using allometry of leaf area, not leaf mass. Optimization models probably already give the right answer because normally LAI is optimized (Woodward).

A recent study compared allocation routines in a number of leading ecosystem models \cite{dekauwe2014}, and recommended to constrain allocation by observed biomass fractions instead of using constant allocation fractions, which is common practice. As a first approximation, which is useful for large scale DGVMs, models should aim to keep leaf area to stem cross-sectional area constant, and vary biomass allocation accordingly. Specific leaf area then gives the ratio of leaf biomass to stem biomass.

There is no shortage of theoretical developments in understanding biomass allocation (\cite{cannelldewar1994,franklin2012,enquist2012}). In particular, theoretical work focusing on optimal partitioning theory is promising (e.g. \cite{valentinemakela2012}) but is yet to lead to operational models for dynamic global vegetation models.

We found no effects of MAP on biomass partitioning, and only very weak effects of MAT on LMF, when analyzed in addition to PFT. This means that climate-related differences are captured by a shift in PFT (see also Fig.~\ref{fig:figure1}, showing PFTs occupy different parts of the climate space). Our results contrast with those of \cite{reich2014} but they studied whole-stand biomass, not individual plant biomass. So perhaps this means the climate effects are on stand density not individual plant effects. No effect of climate is good news for models that parameterize the globe by PFT, as this seems sufficient, based on our dataset.

@@

We used the recently compiled Biomass And Allometry Database (BAAD) (Falster et al., in review), including a total of \Sexpr{nrow(baad_all$data)} observations, but here we use only data that include leaf mass ($M_F$), leaf area ($A_F$), above-ground woody biomass ($M_W$), plant height ($H$), and stem area measured at ground level ($A_S$), or at breast height (ca. 1.3m) ($A_{Sbh}$).  For each analysis, we used different subsets of the data. Sample sizes by PFT are summarized in Tables~\ref{table:count1}-\ref{table:count4} We excluded glasshouse and common garden studies, including only field-grown woody plants (including natural vegetation, unmanaged and managed plantations). We considered three PFTs : evergreen angiosperms, evergreen gymnosperms, and deciduous angiosperms. We excluded  deciduous gymnosperms because few data were available. All locations were further separated into boreal (including sub-boreal), temperate, and tropical biomes.

@@

We only used LMA directly estimated for the harvested plants (typically for a subsample of leaves). For conifers, leaf area was converted to half-total surface area, which is most appropriate for comparison to flat leaves in terms of light interception  \cite{lang1991, chen1992}.

Stem cross-sectional area was measured either at breast height (ca. 1.3m) and/or at the base of the plant. For the subset of the data where both were measured, we estimated basal stem area ($A_S$) from breast height stem area ($A_{Sbh}$) using non-linear regression (see~\ref{sec:predas}). The main results were similar when we repeated it for data where either $A_S$ or $A_{Sbh}$ were directly measured (not shown).

We used a generalized additive model (GAM) to visualize the relationships between biomass and plant size variables. In addition, we corrected for plant size by fitting linear mixed-effects models (with species within study as the random effect) using plant height (linear and squared terms), and calculating least-square means with the 'lsmeans' package in R. All variables were log-transformed prior to analysis. Variance partitioning with marginal (fixed effects) and conditional (fixed and random effects) $R^2$ for linear mixed-effects models as reported by \cite{nakagawa2013}.

To test for direct climate effects in addition to the PFT effect, we used moisture index (rainfall / potential evapotranspiration) and growing season temperature (average air temperature during the period of the year when air temperature is larger than 0) as estimated by Lin et al. (in review). To assess the coverage of the global climate space by the dataset, we extracted mean annual temperature and precipitation from Worldclim \cite{worldclim}, excluding areas without woody vegetation (taken from the global land cover database GLC-SHARE \cite{lathamglobal}).

>>>>

%texcount manuscript.tex > wordcount.txt

@@

\textbf{The partitioning of biomass into short-lived leaves and long-lived stems is one of the most uncertain  and influential components of  global vegetation models (GVMs), affecting the lifetime of carbon in ecosystems \cite{friend2014}, predictions of global carbon storage \cite{ise2010}, and the CO\textsubscript{2} response of net primary production \cite{dekauwe2014}. Although GVMs typically assume that the major woody plant functional types (PFTs) differ in biomass partitioning, empirical data have been lacking. Using a new global database on biomass partitioning from over 175 studies, we here show a strong effect of PFT on partitioning between stems and leaves. Contrary to prior expectation, we show that major woody PFTs do not differ in the amount of leaf area supported, but rather the amount of leaf mass per unit above ground biomass. Moreover, we find that these differences relate to systematic shifts in leaf mass per area, a leading trait within the leaf economics spectrum. Combined, these results uncover fundamental rules in how plants are constructed and allow for systematic benchmarking of biomass partitioning routines in current GVMs.
}

@@

Globally, forests store ca. 360 Pg C in live biomass \cite{panetal2011}, or about 40 years of current antropogenic carbon emissions (Quere). The partioning of this biomass among plant components is a major determinant of vegetation productivity and the rate of carbon cycling. Biomass stored in stems has a long residence time \cite{luyssaert2008},  thus representing a significant long-term carbon storage term. Leaf biomass turns over quickly, and is a major input to the soil carbon cycle where the majority of carbon is released back to the atmosphere. Recently, the residence time of carbon in vegetation has been identified as a major source of uncertainty in model projections of effects of climate change on the global carbon sink \cite{friend2014}. There is little consensus in these global vegetation models (GVMs) on how biomass partitioning should be modelled \cite{dekauwe2014}, but there is no shortage of theory underling the complex process of biomass allocation \cite{franklin2012} (we use allocation to mean the flux of carbon into plant components, and partitioning the actual distribution of biomass among components, see Methods S1).

It is difficult to decide which partioning models and parameterisations are superior, since GVMs are normally tested against productivity or carbon storage data, and both are the outcome of multiple processes. The vast majority of GVMs simplify parameterization by considering a few plant functional types (PFTs). It is also common to assume that biomass partitioning varies between PFTs, however, parameterization of these differences have been based on little empirical evidence.  Based on previous compilations of forest biomass estimates, it is known that gymnosperms have more leaf biomass per unit above-ground biomass (the 'leaf mass fraction', LMF) than angiosperms \cite{oneilldeangelis1981,enquistniklas2002,reich2014}. However, it is not known whether this is due to the angiosperm/evergreen divergence, or related to inherent differences in leaf life span. If the latter is true, we would expect evergreen angiosperms to have more leaf biomass than deciduous angiosperms.

We argue that the lack of a large global database on biomass partitioning at the individual plant level has hampered progress in this field. Here, we use such a database to lay the empirical base against which models can be benchmarked. Our new database (BAAD, Falster) includes over 20,000 observations of biomass and size metrics for individual plants, covering the entire size range including saplings and the largest trees in the world, three major woody PFTs, and three biomes (tropical, temperate, and boreal forests). Another unique aspect is that the database contains leaf mass as well as area estimates, previous compilations only considered mass. It is unclear whether PFTs with higher leaf biomass will have higher total leaf area, because PFTs differ in the ratio of leaf mass to area (LMA), a measure of the leaf construction cost \cite{poorter2009}. Leaf area itself is relevant because light interception acts on leaf area, which drives photosynthesis. A few classic studies have assumed that gymnosperms indeed carry much more leaf area than angiosperms \cite{chabothicks1982, bond1989}. Finally it is also unknown whether potential differences between PFTs in biomass partitioning are a result of inherent differences in morphology, or whether climate directly affects paritioning, and the PFTs simply occupy different climate spaces.

We hypothesize that gymnosperms have higher LMF than angiosperms, and that evergreen angiosperms have higher LMF than deciduous angiosperms (we did not have sufficient data to test deciduous evergreens). We further test whether differences between PFTs are related to LMA. We use several variables from BAAD (Fig. 1a) to test this main hypothesis. We also test whether climate has direct influence on biomass partitioning. This is possible because the database has very good global coverage of the climate space (Fig. 1b).

We found that LMF differed strongly between the three PFTs tested, with evergreen gymnosperms > evergreen angiosperms > deciduous angiosperms (Fig. 2a). Because of allometric relationships between the studied variables \cite{niklas1994}, there was a very strong effect of plant height on LMF (Fig 2a and S2). We corrected for plant height by estimating LMF at a common plant height, using a generalized additive model, a semi-parametric method that makes no assumption on the underlying functional relationships. We then found that LMF at a common plant height was proportional to the leaf construction cost, as measured by leaf mass per area (Fig 2b). The consequence of this is that the leaf area ratio (the amount of leaf area supported per unit above-ground biomass), was invariant between PFTs, because $LAR = LMF / LMA$ (Fig 2c and S3). We thus supported the view that PFTs differ in LMF, but for the first time demonstrated that they converge in the amount of leaf area supported.

Next, we decomposed LMF into two components, $LMF = M_F/M_T = M_F/A_S * A_S/MT$, where $M_F/A_S$ is the ratio of leaf biomass to stem basal area (see Methods S2), and $A_S/M_T$ the ratio of stem basal area to total above-ground biomass. We found that the first term showed identical PFT patterns as LMF (Fig 2d and e), $M_F/A_S$ was highly variable between PFT and proportional to LMA, so that $A_F/A_S$ is invariant between PFTs. This decomposition is also convenient because the first term is nearly independent of plant height, allowing a more straightforward analysis of climate effects (not having to correct for height every time). The second term was found to be nearly unaffected by PFT (Fig. S4), but strongly height-dependent. It thus absorbs variation related to height and above-ground plant geometry, and is consistent with previous findings that this ratio is fairly conserved between plant species \cite{chave2005}.

We analyzed climate effects in two ways: with quantitative variables (mean annual precipitation (MAP), mean annual temperature (MAT)), and by biome (tropical / temperate/ boreal), which is a simple classification taking into account both MAT and MAP. Regardless of how we analyzed it, we found only weak climate effects. We did not find differences between biomes in $M_F/A_S$ (Fig. 3), but some small inconsistent differences in $A_F/A_S$ (corresponding to slightly lower LMA in tropical vs. temperate evergreen angiosperms). Biome consistently explained very little variation when added to a statistical model in addition to PFT and plant height (Table S6). Likewise, we found weak and inconsistent effects of climate variables (MAP and MAT) on $M_F/A_S$ and $A_F/A_S$ (Fig 4). These variables explained very little variation when added to a statistical model in addition to PFT and plant height (Table 1).

We found that PFTs differed in LMF, and this was proportional to leaf mass per area (LMA), a leading trait within the leaf economics spectrum. Indeed, if LMA was used instead of PFT, we were able to explain more variation in $M_F/A_S$ and $A_F/A_S$ (Table S7). It thus seems possible to integrate a key leaf trait with whole-plant modelling of biomass partitioning. One explanation for this proportionality may be the correlation between LMA and leaf lifespan. If annual leaf production is constant, longer leaf lifespan will simply result in the accumulation of more leaf cohorts, and thus biomass \cite{enquistniklas2002}. However, leaf lifespan is correlated to but not proportional to LMA \cite{wrightetal2004}. Another interpretation is that plants may target a certain leaf area display given a certain biomass investment in supporting tissues, because light interception and photosynthesis correlate with the exchange area of the plant.

We found very only weak and inconsistent effects of climate on biomass partitioning, in contrast to a recent study where LMF was found to be correlated with MAT \cite{reich2014}. However, they studied whole-stand biomass, not individual plant biomass. This begs the question whether climate effects on stand-level biomass partitioning are on stand density or individual plants. Our finding that climate does not appreciably affect biomass partitioning is good news for models that parameterize the globe by PFT alone. We do stress that there is still much variation between species and studies unaccounted for (Fig. S5), and that other studies certainly have uncovered small-scale effects of climate on leaf area to sapwood area \cite{callaway1994}, an important component of $A_F/A_S$. Coarse climate variables such as MAT and MAP may mask small-scale climate effects on biomass partitioning.

A recent study compared allocation routines in a number of leading ecosystem models \cite{dekauwe2014}, and recommended to constrain allocation by observed biomass fractions instead of using constant allocation fractions. We argue that these models should be constrained by patterns with plant size and PFT presented in our study. As a first approximation, which is useful for GVMs, models should aim to keep leaf area to stem cross-sectional area constant, and vary biomass allocation accordingly. The leaf mass per area, already a parameter in most GVMs, then gives the ratio of leaf biomass to stem biomass. The data we use here is freely available (Falster), which should aid in the models being tested, with the hope that they perhaps converge.

@@

We used the recently compiled Biomass And Allometry Database (BAAD) (Falster), including a total of \Sexpr{nrow(baad_all$data)} observations, but here we use only data that include leaf mass ($M_F$), leaf area ($A_F$), above-ground woody biomass ($M_W$), plant height ($H$), and stem area measured at ground level ($A_S$), or at breast height (ca. 1.3m) ($A_{Sbh}$).  For each analysis, we used different subsets of the data. Sample sizes by PFT are summarized in Tables~\ref{table:count1}-\ref{table:count4} We excluded glasshouse and common garden studies, including only field-grown woody plants (including natural vegetation, unmanaged and managed plantations). We considered three PFTs : evergreen angiosperms, evergreen gymnosperms, and deciduous angiosperms. We excluded  deciduous gymnosperms because few data were available. All locations were further separated into boreal (including sub-boreal), temperate, and tropical biomes. To assess the coverage of the global climate space by the dataset, we extracted mean annual temperature and precipitation from Worldclim \cite{worldclim}, excluding areas without woody vegetation (taken from the global land cover database GLC-SHARE \cite{lathamglobal}).

@@

We only used LMA directly estimated for the harvested plants (typically for a subsample of leaves). For conifers, leaf area was converted to half-total surface area using published conversion factors \cite{barclay2000}, which is most appropriate for comparison to flat leaves in terms of light interception  \cite{lang1991, chen1992}. Stem cross-sectional area was measured either at breast height (ca. 1.3m) and/or at the base of the plant. For the subset of the data where both were measured, we estimated basal stem area ($A_S$) from breast height stem area ($A_{Sbh}$) using non-linear regression (see~\ref{sec:predas}).

We used generalized additive models (GAM) to visualize the relationships between biomass and plant size variables, and to estimate variables and their confidence intervals such as LMF at a common plant height. In all fitted GAMs, we used species-dataset combination as a random effect. We used the \emph{mgcv} package in R \cite{wood2006}. All variables were log-transformed prior to analysis.

Variance explained by quantitative climate variables (MAP and MAT) were tested with GAMs where variables were sequentially added to the model, and the explained variance ($R^2$) calculated. For the test of biome effects on biomass partitioning, we used a linear mixed-effects model (because two factors and their interactions were tested), again with species-dataset as the random effect. We calculated the $R^2$ for linear mixed-effects models as reported by \cite{nakagawa2013} for the fixed effects only.
